Topic modeling on EU and the USA research projects

### 2017-06-13 thesis_report/draft_thesis.tex
Rewrite VB and online VB LDA

We can analyze a corpus of documents with LDA by examining the posterior
distribution of topics $\beta$, topic proportions $\theta$ and topic
assignments $z$ conditioned on the documents. This reveals latent structure in
the collection that can be used for prediction or data exploration. This
posterior can't be computed directly; it is usually approximated using Markov
Chain Monte Carlo (MCMC) or variational inference. Both methods are effective,
but also present significant computational challenges in large data sets
~\cite{onlineLDAvb}. In our implementation, we're using the online variational
LDA model, which is described in section 2.2.3.4.

@@

Consider a statistical model accompanied by a set of observed data $X$, a set of
latent data $Z$, a set of latent variables $\theta$ and a likelihood function
$L(\theta \mid X, Z)$. The marginal likelihood of the observed data is then

\begin{equation}
L(\theta \mid X) = p(X \mid \theta) = \sum\limits_{Z} p(X, Z \mid \theta)
\end{equation}

which determines the maximum likelihood estimate of the latent variables. The
sum is however often intractable and the EM algorithm solves this problem by
iterating the following two steps:

\begin{itemize}
\item \textbf{Expectation step (E - step)}: For the current estimate of the 
parameters $\theta^t$, calculate the expected value of the log likelihood function:
\begin{equation}
Q(\theta \mid \theta^t) \leftarrow \mathbb{E}_{Z \mid X, \theta^t}
 [\log(L(\theta \mid X, Z))]
\end{equation}
\item \textbf{Maximization step (M - step)} Maximize the expected value function 
with regard to the parameters:
\begin{equation}
\theta^{t + 1} \leftarrow arg \max\limits_{\theta} Q(\theta \mid \theta^t)
\end{equation}
\end{itemize}

@@

Learning the various distributions (the set of topics, their associated word
probabilities, the topic of each word, and the particular topic mixture of each
document) is a problem of Bayesian inference. In the original paper it's being
used a variational Bayes approximation of the posterior distribution;\cite{blei2003latent} alternative inference techniques use Gibbs sampling and
expectation propagation. The goal of variational inference is to approximate an
intractable probability distribution $p$, with a tractable one $q$, in a way
that makes them as close as possible given a similarity measure. In this thesis,
variational inference is used to approximate the posterior distribution of
latent variables. A new distribution over the hidden variables $q(z, \beta)$
(called the variational distribution) is  introduced. This new distribution has
properties such that it can be efficiently computed.\cite{Fox2011ATO}

The variational distribution is a function of a set of free parameters that are optimized such that the variational distribution is as close as possible to the 
actual target posterior distribution where closeness is measured in terms of KL divergence.
Minimizing the KL divergence between the variational distribution and the target
posterior is equivalent to maximizing the Evidence Lower BOund (ELBO) which is

\begin{equation}
L(q) = E_q [log p(x, z, \beta)] -  E_q [log q(z, \beta)]  
\end{equation}

The variational distribution has the property that it can be efficiently
computed. This is done by making each hidden variable independent of each other.

\begin{equation}
q(z, \beta) = q(\beta \mid \lambda) \prod\limits_{n=1}^N \prod\limits_{j=1}^J q(z_{nj} \mid \phi_{nj})   
\end{equation}

The posterior over the per-word topic assignments $z$ is parameterized by
$\phi$, the posterior over per-document topic weights $\theta$ is parameterized
by $\gamma$, and the posterior over the topics $\beta$ is parameterized by
$\lambda$. Furthermore, each hidden variable is governed by its own variational
parameter so the variational parameters $\lambda$ governs the global variables
$\beta$ and the variational parameters $\phi_n$ govern the local variables $z_n
q(\beta \mid \lambda)$ and $q(z_{nj} \mid \phi_{nj})$ take the same form as the
complete conditionals $p(\beta \mid x, z, \alpha)$ and $p(z_{nj} \mid x_n, z_{n,
-j}, \beta)$, but the natural parameters are now $\lambda$ and $\phi_{nj}$,
respectively to give

\begin{equation}
q(\beta \mid \lambda) = h(\beta) exp(\lambda^T t(\beta) - \alpha_g(\lambda))
\end{equation}
\begin{equation}
q(z_{nj} \mid \phi_{nj}) = h(z_{nj}) exp(\phi_{nj}^T t(z_{nj}) - \alpha_l(\phi_{nj}))
\end{equation}

Where $h(.)$ is called the base measure, $\alpha(.)$ is called the log-
normalizer, $\eta(.)$ is called the natural parameter, and $t(.)$ is called the
sufficient statistics. Furthermore, we also assume that the prior distribution
over $\beta$ is part of the exponential family. We maximize the ELBO objective
function with a coordinate ascent procedure. We find its gradient with respect
to the global variational parameter $\lambda$ and find the value of $\lambda$
that sets the gradient to zero. We do the same thing for the local parameters
$\phi_n$. We iterate between these updates until we converge to the maximum of
the ELBO. The updates are given without proof, but the general procedure is to
write the ELBO in terms of parameter of interest (either $\lambda$ or $\phi_n$)
then take the gradient and set it to zero.

\begin{equation}
\lambda = E_q[\eta_g(x, z, \alpha)]
\end{equation}
\begin{equation}
\phi_{nj} = E_q[\eta_l(x_n, z{n, -j}, \beta)]
\end{equation}

Therefore the updates of each variational parameter are equal to the expected
value of the natural parameters of the complete conditionals. A simple outline 
of the algorithm ~\cite{alexminnar} is given below

\begin{algorithm}[H] 
\SetAlgoLined

Initialize $\lambda^{(0)}$ randomly.
\Repeat{the ELBO converges}{
	\For{each local variational parameter $\phi_{nj}$}{\label{forins}
		Update $\phi_{nj}^{(t)} = \mathbb{E}_{q^{(t - 1)}}[\eta_{l,j}(x_n, z_{n,-j}, \beta)]$.
	}
	Update the global variational parameters, $\lambda^{(t)} = \mathbb{E}_{q^{(t)}}[\eta_g (z_{1:N}, x_{1:N})]$.
}
\caption{Variational Bayes for LDA} 
\end{algorithm}

As you can see, in the local parameter update (in the $for$ loop), we have to
iterate over every data point in the data set which is computationally
expensive. The algorithm above has constant memory requirements and empirically
converges faster than collapsed Gibbs sampling. However, it still requires a
full pass through the entire corpus in each iteration. It can be slow to apply
to very large datasets.

@@

The $t$-th document (or equivalently a vector of words counts) we perform the
E-step like Variational inference to find locally optimal values for $\gamma_t$
and $\phi_t$ while holding $]\lambda$ fixed. And then (the M-step) compute the
optimal value for $\lambda$ if the whole document corpus consisted of this
$t$-th document repeated $M$ times (M: the number of documents in the corpus).
After we have the optimal value of $\lambda$ for the   document, $\lambda$*, we
update the value of $\lambda$ using a weighted average of its previous value and
$\lambda_d$∗. The weight given to $\lambda$∗ is given by 
$\rho_t = (\tau_{0} + t)^{-\kappa}$, with $\kappa \in (0.5, 1]$. 
$\kappa$ controls the importance we give to most recents values of $\lambda$∗.

In short, we approximate the true distribution by a simple distribution $q$, and
associate parameters $\phi,\gamma,\lambda$ with the original parameters $z,\theta,
\beta$ respectively. Recall that $z$ gives the topic assignments for each
word in each document, $\theta$ gives the topic composition of each document, 
and $\beta$ gives the word-topic probabilities for each word and each topic.

@@

Algorithm 2 shows the inference process given a vector of word counts. The
explicit formulas for $\mathbb{E}_q[log \theta_{tk}]$ and $\mathbb{E}_q[log
\beta_{kw}]$ are:

\begin{equation}
\mathbb{E}_q[log \theta_{tk}] = \Psi(\gamma_{tk}) - \Psi(\sum_{i=1}^K \gamma_{ti})
\end{equation}

\begin{equation}
\mathbb{E}_q[log \beta_{kw}] = \Psi(\lambda_{kw}) - \Psi(\sum_{i=1}^W \lambda_{ki})
\end{equation}

Where $\Psi$ denotes the derivative of the logarithm of the gamma function.

\begin{algorithm}
\SetAlgoLined
Define $\rho_t = (\tau_{0} + t)^{-\kappa}$

Initialize $\lambda$ randomly.

\For{$t = 0 to \infty$}{\label{forins}
	E step:

	Initialize $\gamma_{tk} = 1$

	\Repeat{$\frac{1}{K} \sum_k$ |change in $\gamma_{tk}$| $< 0.00001$}{
		Set $\phi_{twk} \propto exp${$\mathbb{E}_q[log \theta_{tk}]$ + $\mathbb{E}_q[log\beta_{kw}]$}

		Set $\gamma_{tk} = \alpha + \sum_w \phi_{twk}n_{tw}$
	}
	M step:

	Compute $\lambda_{kw}* = \eta + Dn_{tw}\phi_{twk}$

	Set $\lambda = (1 - \rho_t)\lambda + \rho_{t}\lambda*$
}

We can see that unlike variational LDA, in online LDA we only need to hold a
small chunk of the data at a time, and once we’re done analyzing it, we never
need it again. As with batch, once we’ve estimated $\lambda$, we can find the 
most probable words for each topic by looking at the word probabilities in 
each row of $\lambda$.

\caption{Online Variational Bayes for LDA} 
\end{algorithm}

>>>>

We can see that this does in fact generate a document based on the topic 
mixture $\theta$, the topic-word assignments $z$, and the probability 
matrix $\beta$. Then, we can analyze a corpus of documents with LDA by 
examining $\beta$, which is the posterior distribution of topics, $\theta$ 
and $z$ conditioned on the documents. This reveals latent structure in
the collection that can be used for prediction or data exploration.

Hence, we observe the document, and must infer the latent topic mixture 
$\theta$ and topic-word assignments $z$. LDA aims to infer:

\begin{equation}
p(\theta, \textbf{z} |\textbf{w}, \alpha, \beta)
\end{equation}

This
posterior can't be computed directly; it is usually approximated using Markov
Chain Monte Carlo (MCMC) or variational inference. Both methods are effective,
but also present significant computational challenges in large data sets
~\cite{onlineLDAvb}. In our implementation, we're using the online variational
LDA model, which is described in section 2.2.3.4.

@@

The general idea behind the Expectation-Maximization algorithm is this:

\begin{enumerate}
\item Start with an initial estimate of what each parameter might be.
\item \textbf{E step}: Compute the likelihood that each parameter produces 
the data point.
\item Calculate weights for each data point based on the likelihood of it 
being produced by a parameter.
\item \textbf{M step}: Combine these weights together with the data to 
compute a better estimate for the parameters.
\item Repeat steps 2 to 4 until the parameter estimate converges.
\end{enumerate}

@@

In LDA, our ultimate goal is to find the posterior distribution of latent topic variables after observing training data. However, computing this distribution is intractable, so we’re forced to approximate. One approximation approach is to 
use an optimization method called Variational Bayes. In the original paper it's 
being used a variational Bayes approximation of the posterior distribution;\cite{blei2003latent} alternative inference techniques use Gibbs sampling and
expectation propagation. The goal of variational inference is to approximate an
intractable probability distribution $p$, with a tractable one $q$, in a way
that makes them as close as possible given a similarity measure. In this thesis,
variational inference is used to approximate the posterior distribution of
latent variables. A new distribution over the hidden variables $q(z, \beta)$
(called the variational distribution) is  introduced. This new distribution has
properties such that it can be efficiently computed.\cite{Fox2011ATO}

In short, we approximate the true distribution by a simple distribution $q$, and associate parameters $\phi$, $\gamma$, $\lambda$ with the original parameters 
$z$, $\theta$, $\beta$ respectively. Recall that $z$ gives the topic assignments 
for each word in each document, $\theta$ gives the topic composition of each 
document, and $\beta$ gives the word-topic probabilities for each word and each 
topic. The variational distribution is a function of a set of free parameters that 
are optimized such that the variational distribution is as close as possible to 
the actual target posterior distribution where closeness is measured in terms of 
KL divergence.

Specifically, we have:
\begin{equation}
q(z_{di}=k) = \phi_{dw_{di}k}
\end{equation}
\begin{equation}
q(\theta_{d}) = dirichlet(\theta_{d}; \gamma_{d})
\end{equation}
\begin{equation}
q(\beta_{k}) = dirichlet(\beta_{k}; \lambda_{k})
\end{equation}

The posterior over the per-word topic assignments $z$ is parameterized by
$\phi$, the posterior over per-document topic weights $\theta$ is parameterized
by $\gamma$, and the posterior over the topics $\beta$ is parameterized by
$\lambda$.

Our goal is to estimate $\phi$, $\gamma$ and $\lambda$. In both VB and 
online VB LDA, we alternate between two steps:
\begin{enumerate}
\item E-Step: Estimate $\phi$, $\gamma$ using the current value of $\lambda$
\item M-Step: Update $\lambda$, using the current value of $\phi$
\end{enumerate}

In Variational Bayes, we perform multiple passes over the entire dataset, 
checking each time for convergence. During each pass, the algorithm does an 
E-Step using the entire dataset and an M-Step, which updates $\lambda$ using 
$\phi$ values from every document. At a high level:

\begin{algorithm}
\SetAlgoNoLine
E Step:

\For{$d = 1$ to $numDocs$}{
	initialize $\gamma$

    \Repeat{change in $\phi < \epsilon$}{
        update $\phi$
        update $\gamma$
     }
}
M Step:

update $\lambda$
\end{algorithm}

The specific updates are:
\begin{equation}
\phi_{d,word_{i},k} \propto e^{E_{q}(log \theta_{d,k})+E_{q}(log\beta_{k,word_{i}})}
\end{equation}
\begin{equation}
\gamma_{d,k} = \alpha + \sum_{word_{i}}\phi_{d,word_{i},k}n_{d,word_{i}}
\end{equation}
\begin{equation}
\lambda_{k,word_{i}}=\eta +\sum_{d}n_{d,word_{i}}\phi_{d,word_{i},k}
\end{equation}
Where $n_{d,word_{i}}$ is the number of occurrences of $word_{i}$ in document $d$.
The algorithm above has constant memory requirements and empirically
converges faster than collapsed Gibbs sampling. However, it still requires a
full pass through the entire corpus in each iteration. It can be slow to apply
to very large datasets.

@@

In short, we approximate the true distribution by a simple distribution $q$, and
associate parameters $\phi,\gamma,\lambda$ with the original parameters $z,\theta,
\beta$ respectively, as in VB. Recall that $z$ gives the topic assignments for each
word in each document, $\theta$ gives the topic composition of each document, 
and $\beta$ gives the word-topic probabilities for each word and each topic.

@@

In online Variational Bayes, we only make a single sweep of the entire dataset, analyzing a chunk of documents at a time. A chunk could be a single document, 
50 documents, or even the entire dataset. Let’s let a chunk be 1000 documents.

The online E-Step only uses the current chunk; instead of holding all the 
documents we now only have to hold 1000 in memory. The E-Step finds locally 
optimal values for $\phi$ and $\gamma$, while in the M-Step, we first compute 
$\lambda^{*}$, which is the value of $\lambda$ if we imagined that the entire 
dataset is made up of $\frac{numDocs}{chunkSize}$ copies of the current chunk. 
Then $\lambda$ is updated using a weighted sum of $\lambda^{*}$ and $\lambda$:

\begin{algorithm}
\SetAlgoNoLine
E Step:

initialize $\gamma$

\Repeat{change in $\phi < \epsilon$}{
        update $\phi$

        update $\gamma$
}
M Step:

compute $\lambda^{*}$

update $\lambda$
\end{algorithm}

We can see that unlike variational LDA, in online LDA we only need to hold a
small chunk of the data at a time, and once we’re done analyzing it, we never
need it again. As with batch, once we’ve estimated $\lambda$, we can find the 
most probable words for each topic by looking at the word probabilities in 
each row of $\lambda$.\footnote{\url{https://wellecks.wordpress.com/2014/10/26/ldaoverflow-with-online-lda/}}
